---
title: "Investment-1"
author: "Geirmundur Orri Sigurdsson"
date: '2023-01-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

```{r dependencies, message=FALSE, warning=FALSE}
library(quantmod)
library(tidyverse)
library(lubridate)
library(knitr)
```

# investment Data Frame

```{r investment}
raw <- read.csv('../../tmp/finance/investment.csv')
investment <- raw %>%
  mutate(Date = as_datetime(Date)) %>%
  select(Date, Symbol, Shares, Price)
```

# Fetch Current Price

```{r fetch}
price_table <- hashtab()
voo_quote <- getQuote('VOO')
qqq_quote <- getQuote('QQQ')
amzn_quote <- getQuote('AMZN')
goog_quote <- getQuote('GOOG')
sethash(price_table, 'VOO', voo_quote$Last)
sethash(price_table, 'QQQ', qqq_quote$Last)
sethash(price_table, 'AMZN', amzn_quote$Last)
sethash(price_table, 'GOOG', goog_quote$Last)
```

# Calculate

```{r calculate}
today <- as_datetime(Sys.Date())
holdings <- investment %>% mutate(Original = Price * Shares)
for (row in 1:nrow(holdings)) {
  holdings[row, 'Current'] = price_table[[holdings$Symbol[row]]]
}
holdings <- holdings %>% mutate(Value = Current * Shares)
holdings <- holdings %>% mutate(Profit = Value - Original)
holdings <- holdings %>% mutate(Ratio = Profit / Original, Elapsed = today - Date)
holdings <- holdings %>% mutate(Years = as.numeric(as.duration(Elapsed)) / 31536000)
holdings <- holdings %>% mutate(PPY = ifelse(Years > 1.0, Profit / Years, Profit))
holdings <- holdings %>% mutate(PPYR = PPY / Original)
total_buy <- sum(holdings$Original)
total_fee <- sum(holdings$Fee)
total_profit <- sum(holdings$Profit)
profit_ratio <- total_profit / total_buy
total_ppy <- sum(holdings$PPY)
ppyr <- total_ppy / total_buy
rm(row, today)
```

# Output Table

```{r output}
kable(holdings, caption = 'Holding')
```

# Visualisation

```{r visualisation, warning=FALSE}
ggplot(data = holdings, mapping = aes(x = Date, y = Profit, color = Symbol)) + geom_line() + theme_light()
```

# Results

Totally invested is $`r toString(total_buy)`. Total profit is $`r toString(round(total_profit, digits = 0))`. Profit ratio is `r toString(round(100*profit_ratio, digits = 2))`%. Profit per year is $`r toString(round(total_ppy, digits = 0))`. Profit per year ratio is `r toString(round(100*ppyr, digits = 2))`%. Amount after selling is $`r toString(total_buy + round(total_profit, digits = 0))`.
