---
title: "Investment-1"
author: "Geirmundur Orri Sigurdsson"
date: '2023-01-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

```{r dependencies, message=FALSE, warning=FALSE}
library(quantmod)
library(tidyverse)
library(lubridate)
library(googlesheets4)
library(knitr)
```

# investment Data Frame

```{r investment}
# Suspend any previous Google authorization
gs4_deauth()
# Import from Google Sheets
raw <- read_sheet('https://docs.google.com/spreadsheets/d/1pMQ1PUSKw4Fd7f21yGokdLx1EuQtn2KAGvaGyrjNU7Q')
investment <- raw %>%
  mutate(Date = as_datetime(Date)) %>%
  select(Date, Symbol, Shares, Price)
```

# Fetch Current Price

```{r fetch}
price_table <- hashtab()
for (row in 1:nrow(investment)) {
  symbol <- investment$Symbol[row]
  quote <- getQuote(symbol)
  sethash(price_table, symbol, quote$Last)
}
```

# Calculate

```{r calculate}
today <- as_datetime(Sys.Date())
holdings <- investment %>% mutate(Investment = Price * Shares)
for (row in 1:nrow(holdings)) {
  holdings[row, 'Last'] = price_table[[holdings$Symbol[row]]]
}
holdings <- holdings %>% mutate(Worth = Last * Shares)
holdings <- holdings %>% mutate(Profit = Worth - Investment)
holdings <- holdings %>% mutate(Ratio = Profit / Investment, Elapsed = today - Date)
holdings <- holdings %>% mutate(Years = as.numeric(as.duration(Elapsed)) / 31536000)
holdings <- holdings %>% mutate(PPY = ifelse(Years > 1.0, Profit / Years, Profit))
holdings <- holdings %>% mutate(PPYR = PPY / Investment)
total_inv <- sum(holdings$Investment)
total_worth <- sum(holdings$Worth)
total_profit <- sum(holdings$Profit)
profit_ratio <- total_profit / total_inv
total_ppy <- sum(holdings$PPY)
ppyr <- total_ppy / total_inv
holdings <- holdings %>% mutate(IR = Investment / total_inv, WR = Worth / total_worth)
rm(row, today)

result <- data.frame(
  Result = c(
    "Total Investment",
    "Total Worth",
    "Total Profit",
    "Profit Ratio"),
  Value = c(
    total_inv,
    total_worth,
    total_profit,
    profit_ratio * 100.0),
  Unit = c(
    "USD",
    "USD",
    "USD",
    "%"))
```

# Output Holding Table

```{r output holding table}
kable(holdings, caption = 'Holding', digits = 2)
```

# Visualisation

```{r visualisation, warning=FALSE}
ggplot(data = holdings, mapping = aes(x = Date, y = Investment, fill = Symbol)) + geom_bar(stat = "identity") + theme_light()
ggplot(data = holdings, mapping = aes(x = Date, y = Worth, fill = Symbol)) + geom_bar(stat = "identity") + theme_light()
ggplot(data = holdings, mapping = aes(x = Date, y = Profit, fill = Symbol)) + geom_bar(stat = "identity") + theme_light()
```

# Results

```{r output result table}
kable(result, caption = 'Result', digits = 2)
```

Totally invested is $`r toString(total_inv)`. Total profit is $`r toString(round(total_profit, digits = 0))`. Profit ratio is `r toString(round(100*profit_ratio, digits = 2))`%. Profit per year is $`r toString(round(total_ppy, digits = 0))`. Profit per year ratio is `r toString(round(100*ppyr, digits = 2))`%. Amount after selling is $`r toString(total_inv + round(total_profit, digits = 0))`.
