---
title: 'Investment Simulation #2'
author: "Geirmundur Orri Sigurdsson"
date: '2023-01-25'
output: html_document
params:
  end_date: now
  year_count: 4
  buy_per_year: 12
  voo_buy_budget: 600
  qqq_buy_budget: 400
  amzn_buy_budget: 100
  goog_buy_budget: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

```{r dependencies, message=FALSE, warning=FALSE}
library(quantmod)
library(tidyverse)
library(lubridate)
library(xts)
```

# Option

```{r option}
# Buy count
buy_count = params$year_count * params$buy_per_year

# Data Range
edate <- as.Date(params$end_date, optional = TRUE)
if (is.na(edate)) { edate <- Sys.Date() }
sdate <- edate - years(params$year_count)
```

The Investment Simulation simulates `r toString(params$buy_per_year)` investments per year for `r toString(params$year_count)` years between `r toString(sdate)` to `r toString(edate)`

# Fetch Data

```{r fetch}
voo_stock <- getSymbols('VOO', from=sdate, to=edate, auto.assign=F)
qqq_stock <- getSymbols('QQQ', from=sdate, to=edate, auto.assign=F)
amzn_stock <- getSymbols('AMZN', from=sdate, to=edate, auto.assign=F)
goog_stock <- getSymbols('GOOG', from=sdate, to=edate, auto.assign=F)
```

# Wrangling

```{r wrangling}
# Typically use previous value for NA
no.na <- which(is.na(voo_stock[,6]))
voo_stock[no.na,6] <- voo_stock[no.na-1,6]
no.na <- which(is.na(qqq_stock[,6]))
qqq_stock[no.na,6] <- qqq_stock[no.na-1,6]
no.na <- which(is.na(amzn_stock[,6]))
amzn_stock[no.na,6] <- amzn_stock[no.na-1,6]
no.na <- which(is.na(goog_stock[,6]))
goog_stock[no.na,6] <- goog_stock[no.na-1,6]

# Only stock price
voo_price <- voo_stock[,6]
qqq_price <- qqq_stock[,6]
amzn_price <- amzn_stock[,6]
goog_price <- goog_stock[,6]

voo <- fortify.zoo(voo_price)
voo <- voo %>% mutate(date=Index, tag='VOO', value=VOO.Adjusted) %>% select(date, tag, value)
qqq <- fortify.zoo(qqq_price)
qqq <- qqq %>% mutate(date=Index, tag='QQQ', value=QQQ.Adjusted) %>% select(date, tag, value)
amzn <- fortify.zoo(amzn_price)
amzn <- amzn %>% mutate(date=Index, tag='AMZN', value=AMZN.Adjusted) %>% select(date, tag, value)
goog <- fortify.zoo(goog_price)
goog <- goog %>% mutate(date=Index, tag='GOOG', value=GOOG.Adjusted) %>% select(date, tag, value)
```

# Checking and Extracting

```{r checking}
count <- nrow(voo)
stopifnot(count == nrow(qqq))
stopifnot(count == nrow(amzn))
stopifnot(count == nrow(amzn))

buy_space <- count / buy_count

voo_last <- voo$value[nrow(voo)]
qqq_last <- qqq$value[nrow(qqq)]
amzn_last <- amzn$value[nrow(amzn)]
goog_last <- goog$value[nrow(goog)]
```

# Simulate

```{r simulate}
index <- 1
total_buy <- 0
buy <- data.frame(date=as.Date(character()), tag=character(), price=as.numeric(character()), shares=as.numeric(character()), sell=as.numeric(character()), profit=as.numeric(character()))
while(index < count) {
  by <- nrow(buy) + 1
  buy[by, 1] = voo$date[index]
  buy[by, 2] = voo$tag[index]
  buy[by, 3] = voo$value[index]
  buy[by, 4] = params$voo_buy_budget / voo$value[index]
  buy[by, 5] = voo_last * params$voo_buy_budget / voo$value[index]
  buy[by, 6] = (voo_last * params$voo_buy_budget / voo$value[index]) - params$voo_buy_budget
  total_buy <- total_buy + params$voo_buy_budget
  by <- nrow(buy) + 1
  buy[by, 1] = qqq$date[index]
  buy[by, 2] = qqq$tag[index]
  buy[by, 3] = qqq$value[index]
  buy[by, 4] = params$qqq_buy_budget / qqq$value[index]
  buy[by, 5] = qqq_last * params$qqq_buy_budget / qqq$value[index]
  buy[by, 6] = (qqq_last * params$qqq_buy_budget / qqq$value[index]) - params$qqq_buy_budget
  total_buy <- total_buy + params$qqq_buy_budget
  by <- nrow(buy) + 1
  buy[by, 1] = amzn$date[index]
  buy[by, 2] = amzn$tag[index]
  buy[by, 3] = amzn$value[index]
  buy[by, 4] = params$amzn_buy_budget / amzn$value[index]
  buy[by, 5] = amzn_last * params$amzn_buy_budget / amzn$value[index]
  buy[by, 6] = (amzn_last * params$amzn_buy_budget / amzn$value[index]) - params$amzn_buy_budget
  total_buy <- total_buy + params$amzn_buy_budget
  by <- nrow(buy) + 1
  buy[by, 1] = goog$date[index]
  buy[by, 2] = goog$tag[index]
  buy[by, 3] = goog$value[index]
  buy[by, 4] = params$goog_buy_budget / goog$value[index]
  buy[by, 5] = goog_last * params$goog_buy_budget / goog$value[index]
  buy[by, 6] = (goog_last * params$goog_buy_budget / goog$value[index]) - params$goog_buy_budget
  total_buy <- total_buy + params$goog_buy_budget
  index <- index + buy_space
}
total_profit <- sum(buy$profit)
total_profit_per_year <- total_profit / params$year_count
profit_ratio <- total_profit / total_buy
profit_ratio_per_year <- total_profit_per_year / total_buy
```

# Visualisation

```{r visualisation, warning=FALSE}
ggplot(data = buy, mapping = aes(x = date, y = profit, color = tag)) + geom_line() + theme_light()
```

# Results

Totally invested is $`r toString(total_buy)`. Total profit is $`r toString(round(total_profit, digits = 0))`. Profit ratio is `r toString(round(100*profit_ratio, digits = 2))`%. Profit per year is $`r toString(round(total_profit_per_year, digits = 0))`. Profit per year ratio is `r toString(round(100*profit_ratio_per_year, digits = 2))`%. Amount after selling is $`r toString(total_buy + round(total_profit, digits = 0))`.
