---
title: "Radiation Intensity"
author: "Geirmundur Orri Sigurdsson"
date: "4/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Radiation Intensity

Inverse-square law for light and other electromagnetic radiation

See [Inverse-square law](https://en.wikipedia.org/wiki/Inverse-square_law) on Wikipedia

Formula: $$I = \frac{P}{A} = \frac{P}{4\pi r^{^{2}}}$$

Where I is the intensity or power per unit area

Given an antena with radius r gives area A as: $$A = \pi r^{^{2}}$$

Naming with transmitter, antena and solving for power of the transmitter gives: $$P_{transmitter} = 4\pi d^{^{2}}\times \frac{P_{received}}{A_{antena}}$$

where d is the distance

# Formulating as function

```{r function}
radiation_intensity <- function(distance, Preceived, Rantena) {
  Aantena <- pi * Rantena * Rantena
  4 * pi * distance * distance * Preceived / Aantena
}
```

# Proxima Centauri

Distance to Proxima Centauri is 4.2465 ly and one ly is 9.46E12 km

See [Proxima Centauri](https://en.wikipedia.org/wiki/Proxima_Centauri) on Wikipedia

```{r distance to Proxima Centauri}
Dpc <- 4.2465 * 9.46E12 * 1000 # to meters
```

# Example

1. how powerful does a radio transmitter on Proxima Centauri need to be to detect a signal of 1W with a radio telescope with radius of 100 meter

```{r example 1}
radiation_intensity(Dpc, 1, 100) / 1E15 # in Petawatt
```

2. how powerful does a radio transmitter on Proxima Centauri need to be to detect a signal of 0.1W with a radio telescope with radius of 100 meter

```{r example 2}
radiation_intensity(Dpc, 0.1, 100) / 1E15 # in Petawatt
```

3. how powerful does a radio transmitter on Proxima Centauri need to be to detect a signal of 10W with the FAST a radio telescope with illuminated radius of 150 meter

```{r example 3}
radiation_intensity(Dpc, 10, 150) / 1E15 # in Petawatt
```