---
title: Investigating NASA Pioneer 10 Charged Particle Instrument with distance from
  Sun
author: "Geirmundur Orri Sigurdsson"
date: "4/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

```{r dependencies, message=FALSE, warning=FALSE}
library(psych)      # For describe
library(lubridate)
library(data.table) # For fread
library(sphereplot)
library(tidyverse)
library(foreach)
library(plotly)
```

# Functions

## Function for aggregation

```{r function for aggregation}
aggnrmean <- function(df, n = 5, FUN = mean) {
  aggregate(df,
            by = list(gl(ceiling(nrow(df)/n), n)[1:nrow(df)]),
            FUN = FUN)[-1]
}
```

# Download Data

## Downloading all 1 hour CPI data and combine into one

```{r download and combine cpi, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
path <- 'https://spdf.gsfc.nasa.gov/pub/data/pioneer/pioneer10/particle/cpi/ip_1hour_ascii/p10cp_hr'
yseq <- seq(1972, 1992)
rcpi <- data.frame()      # Empty data frame for result
foreach(y = yseq) %do% {
  url <- paste(path, as.character(y), '.asc', sep = '')
  ycpi <- fread(url)
  rcpi <- rbind(rcpi, ycpi)
}
```

## Download mission Trajectory

```{r download mission trajectory, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
rtraj <- fread("https://spdf.gsfc.nasa.gov/pub/data/pioneer/pioneer10/traj/ip_project/p10tjall.asc")
```

# Convert date time

```{r convert time}
cpi <- rcpi %>% mutate(ts = date_decimal(V1) + days(V2) + hours(V3))
traj <- rtraj %>% mutate(ts = date_decimal(V1) + as.duration(86400 * V2))
```

# Select and rename

```{r select and rename}
cpi <- cpi %>% select(
  ts,
  RID2P   = V4,   # ID-2 rate for 11-20 MeV protons [cps]
  RID2HE  = V5,   # ID-2 rate for 11-20 MeV/nucleon helium [cps]
  RID3P   = V6,   # ID-3 rate for 20-24 MeV protons [cps]
  RID3HE  = V7,   # ID-3 rate for 20-24 MeV/nucleon helium [cps]
  RID4P   = V8,   # ID-4 rate for 24-29 MeV protons [cps]
  RID4HE  = V9,   # ID-4 rate for 24-29 MeV/nucleon helium [cps]
  RID5P   = V10,  # ID-5 rate for 29-67 MeV protons [cps]
  RID5HE  = V11,  # ID-5 rate for 29-67 MeV/nucleon helium [cps]
  RID5E1  = V12,  # ID-5 rate for 7-17 MeV electrons [cps]
  RID5E2  = V13,  # ID-5 rate for 2 x minimum-ionizing [cps]
  RID7    = V14,  # ID-7 + ID-13 integral rate for ions at E > 67 MeV/nucleon [cps]
  RID7ZG5 = V15)  # ID-7 integral rate for Z > 5 ions at E > 67 MeV/nucleon [cps]
traj <- traj %>% select(
  ts,
  HRANGP = V3,  # Distance from Sun to spacecraft in km
  SECLAT = V4,  # Solar ecliptic latitude and
  SECLON = V5,  # longitude of spacecraft with respect to true-of-date Ecliptic
  HELLAT = V6,  # Heliographic latitude and
  HELLON = V7,  # longitude of spacecraft
  HILLON = V8,  # Heliographic inertial longitude of spacecraft with respect to
                # direction of zero heliographic longitude on 1 Jan. 1854 at 1200 UT
  REARSC = V9)  # Distance from Earth to spacecraft in AU
                # (used to calculate Earth Received Time - UT)
```

# Replace fill in value with NA

```{r replace fill with na}
cpi <- na_if(cpi, 1e31)
```

# Time range

## CPI data range

```{r time range cpi}
summary(cpi$ts)
```

## Trajectory data range

```{r time range trajectory}
summary(traj$ts)
```
## Determine Date Range

```{r determine date range}
cpimints <- min(cpi$ts)
cpimaxts <- max(cpi$ts)
trajmints <- min(traj$ts)
trajmaxts <- max(traj$ts)
mints <- max(cpimints, trajmints)
maxts <- min(cpimaxts, trajmaxts)
```

# Filter all data by date range

```{r filter by date range}
cpi <- cpi %>% filter(ts >= mints & ts <= maxts)
traj <- traj %>% filter(ts >= mints & ts <= maxts)
```

# Describe data

```{r describe}
describe(cpi[,-1], skew = FALSE)
```

# Calculate averages

```{r calculate averages}
cpi <- cpi %>% mutate(
  RIDMP = (RID2P + RID3P + RID4P + RID5P) / 4,
  RIDMHE = (RID2HE + RID3HE + RID4HE + RID5HE) / 4,
  RIDME = (RID5E1 + RID5E2) / 2,
  RIDM7 = (RID7 + RID7ZG5) / 2)
```

# Aggregate by Mean

```{r aggregate by mean}
mcpi <- aggnrmean(cpi, 10)
```

## Describe aggregated data

```{r aggregated}
describe(mcpi[,-1], skew = FALSE)
```

# Transforms 3D spherical coordinates to Cartesian coordinates

```{r transforms and combine}
hcart <- sph2car(traj$SECLON, traj$SECLAT, traj$HRANGP)
traj <- traj %>% mutate(x = hcart[,1], y = hcart[,2], z = hcart[,3])
```

# Convert to data tables

```{r convert to data tables}
setDT(mcpi)
setDT(traj)
```

# Combine data

## Combine CPI and Trajectory data into one set

```{r combine v1 and v2 trajectory}
setkey(mcpi, ts)
setkey(traj, ts)
cpitraj <- traj[mcpi, roll = "nearest"]
```

# Plots

## Distance from sun

```{r plots distance, warning=FALSE}
ggplot(data = traj) +
  geom_point(aes(x = ts, y = HRANGP)) +
  labs(title = "Pioneer 10 Trajectory", x = "Time", y = "Distance from Sun (AU)") +
  theme_light()
```

## Protons

```{r plots protons, warning=FALSE}
pcpi <- reshape2::melt(mcpi %>% select(time = ts, "11-20 MeV" = RID2P, "20-24 MeV" = RID3P, "24-29 MeV" = RID4P, "29-67 MeV" = RID5P), id = "time") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = time, y = value, color = variable)) +
  geom_point(aes(x = time, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI Protons", x = "Time", y = "Rate (cps)") +
  theme_light()
pcpi <- reshape2::melt(cpitraj %>% select(distance = HRANGP, "11-20 MeV" = RID2P, "20-24 MeV" = RID3P, "24-29 MeV" = RID4P, "29-67 MeV" = RID5P), id = "distance") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = distance, y = value, color = variable)) +
  geom_point(aes(x = distance, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI Protons", x = "Distance from Sun (AU)", y = "Rate (cps)") +
  theme_light()
```

## Nucleon helium

```{r plots he, warning=FALSE}
pcpi <- reshape2::melt(mcpi %>% select(time = ts, "11-20 MeV" = RID2HE, "20-24 MeV" = RID3HE, "24-29 MeV" = RID4HE, "29-67 MeV" = RID5HE), id = "time") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = time, y = value, color = variable)) +
  geom_point(aes(x = time, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI nucleon He", x = "Time", y = "Rate (cps)") +
  theme_light()
pcpi <- reshape2::melt(cpitraj %>% select(distance = HRANGP, "11-20 MeV" = RID2HE, "20-24 MeV" = RID3HE, "24-29 MeV" = RID4HE, "29-67 MeV" = RID5HE), id = "distance") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = distance, y = value, color = variable)) +
  geom_point(aes(x = distance, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI nucleon He", x = "Distance from Sun (AU)", y = "Rate (cps)") +
  theme_light()
```

## Electrons and minimum-ionizing rate

```{r plots electrons and minimum ionizing, warning=FALSE}
pcpi <- reshape2::melt(mcpi %>% select(time = ts, "7-17 MeV Electrons" = RID5E1, "Double minimum-ionizing" = RID5E2), id = "time") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = time, y = value, color = variable)) +
  geom_point(aes(x = time, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI Electrons and minimum ionizing", x = "Time", y = "Rate (cps)") +
  theme_light()
pcpi <- reshape2::melt(cpitraj %>% select(distance = HRANGP, "7-17 MeV Electrons" = RID5E1, "Double minimum-ionizing" = RID5E2), id = "distance") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = distance, y = value, color = variable)) +
  geom_point(aes(x = distance, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI Electrons and minimum ionizing", x = "Distance from Sun (AU)", y = "Rate (cps)") +
  theme_light()
```

## Integral rate

```{r plots integral, warning=FALSE}
pcpi <- reshape2::melt(mcpi %>% select(time = ts, "Integral rate" = RID7, "Integral rate Z > 5" = RID7ZG5), id = "time") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = time, y = value, color = variable)) +
  geom_point(aes(x = time, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI ions at E > 67 MeV/nucleon", x = "Time", y = "Rate (cps)") +
  theme_light()
pcpi <- reshape2::melt(cpitraj %>% select(distance = HRANGP, "Integral rate" = RID7, "Integral rate Z > 5" = RID7ZG5), id = "distance") %>% filter(!is.na(value))
ggplot(data = pcpi) +
# geom_line(aes(x = distance, y = value, color = variable)) +
  geom_point(aes(x = distance, y = value, color = variable)) +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Pioneer 10 CPI ions at E > 67 MeV/nucleon", x = "Distance from Sun (AU)", y = "Rate (cps)") +
  theme_light()
```

## 3D Scatter

### Average protons rate in Trajectory

```{r figure 3d protons}
p3d <- cpitraj%>% select(x, y, z, Rate = RIDMP)
plot_ly(p3d, x = ~x, y = ~y, z = ~z, color = ~Rate) %>%
  add_markers() %>%
  layout(scene = list(
    xaxis = list(title = 'AU'),
    yaxis = list(title = 'AU'),
    zaxis = list(title = 'AU')))
```

### Average rate of nucleon helium in Trajectory

```{r figure 3d nucleon helium}
p3d <- cpitraj%>% select(x, y, z, Rate = RIDMHE)
plot_ly(p3d, x = ~x, y = ~y, z = ~z, color = ~Rate) %>%
  add_markers() %>%
  layout(scene = list(
    xaxis = list(title = 'AU'),
    yaxis = list(title = 'AU'),
    zaxis = list(title = 'AU')))
```

### Average rate of electrons and ionizing in Trajectory

```{r figure 3d electrons and ionizing}
p3d <- cpitraj%>% select(x, y, z, Rate = RIDME)
plot_ly(p3d, x = ~x, y = ~y, z = ~z, color = ~Rate) %>%
  add_markers() %>%
  layout(scene = list(
    xaxis = list(title = 'AU'),
    yaxis = list(title = 'AU'),
    zaxis = list(title = 'AU')))
```

### Average integral rate in Trajectory

```{r figure 3d integral}
p3d <- cpitraj%>% select(x, y, z, Rate = RIDM7)
plot_ly(p3d, x = ~x, y = ~y, z = ~z, color = ~Rate) %>%
  add_markers() %>%
  layout(scene = list(
    xaxis = list(title = 'AU'),
    yaxis = list(title = 'AU'),
    zaxis = list(title = 'AU')))
```