---
title: "Chapter 7 Exploratory Data Analysis Notebook"
output: html_notebook
---

# R for Data Science

## Chapter 7 Exploratory Data Analysis

### 7.1.1 Prerequisites

```{r prerequisites}
library(tidyverse)
```

### 7.3.1 Visualising distributions

How you visualise the distribution of a variable will depend on whether the variable is categorical or continuous. A variable is categorical if it can only take one of a small set of values. In R, categorical variables are usually saved as factors or character vectors. To examine the distribution of a categorical variable, use a bar chart:

```{r distributions 1}

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))


```

The height of the bars displays how many observations occurred with each x value. You can compute these values manually with dplyr::count():

```{r distributions 2}

diamonds %>% 
  count(cut)

```

A variable is continuous if it can take any of an infinite set of ordered values. Numbers and date-times are two examples of continuous variables. To examine the distribution of a continuous variable, use a histogram:

```{r distributions 3}

ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)

```

You can compute this by hand by combining dplyr::count() and ggplot2::cut_width():

```{r distributions 4}

diamonds %>% 
  count(cut_width(carat, 0.5))

```

You can set the width of the intervals in a histogram with the binwidth argument, which is measured in the units of the x variable. You should always explore a variety of binwidths when working with histograms, as different binwidths can reveal different patterns. For example, here is how the graph above looks when we zoom into just the diamonds with a size of less than three carats and choose a smaller binwidth.

```{r distributions 5}

smaller <- diamonds %>% 
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

```

If you wish to overlay multiple histograms in the same plot, I recommend using geom_freqpoly() instead of geom_histogram(). geom_freqpoly() performs the same calculation as geom_histogram(), but instead of displaying the counts with bars, uses lines instead. It’s much easier to understand overlapping lines than bars.

```{r distributions 6}

ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)

```

### 7.3.2 Typical values

In both bar charts and histograms, tall bars show the common values of a variable, and shorter bars show less-common values. Places that do not have bars reveal values that were not seen in your data. To turn this information into useful questions, look for anything unexpected:

```{r typical 1}

ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

```

The histogram below shows the length (in minutes) of 272 eruptions of the Old Faithful Geyser in Yellowstone National Park. Eruption times appear to be clustered into two groups: there are short eruptions (of around 2 minutes) and long eruptions (4-5 minutes), but little in between.

```{r typical 2}

ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)

```

### 7.3.3 Unusual values

Outliers are observations that are unusual; data points that don’t seem to fit the pattern. Sometimes outliers are data entry errors; other times outliers suggest important new science. When you have a lot of data, outliers are sometimes difficult to see in a histogram. For example, take the distribution of the y variable from the diamonds dataset. The only evidence of outliers is the unusually wide limits on the x-axis.

```{r unusual 1}

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)

```

There are so many observations in the common bins that the rare bins are so short that you can’t see them (although maybe if you stare intently at 0 you’ll spot something). To make it easy to see the unusual values, we need to zoom to small values of the y-axis with coord_cartesian():


```{r unusual 2}

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))

```

(coord_cartesian() also has an xlim() argument for when you need to zoom into the x-axis. ggplot2 also has xlim() and ylim() functions that work slightly differently: they throw away the data outside the limits.)

This allows us to see that there are three unusual values: 0, ~30, and ~60. We pluck them out with dplyr:

```{r unusual 3}

unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)
unusual

```

### 7.4 Missing values

If you’ve encountered unusual values in your dataset, and simply want to move on to the rest of your analysis, you have two options.

Drop the entire row with the strange values:

```{r missing 1}

diamonds2 <- diamonds %>% 
  filter(between(y, 3, 20))

```

Instead, I recommend replacing the unusual values with missing values. The easiest way to do this is to use mutate() to replace the variable with a modified copy. You can use the ifelse() function to replace unusual values with NA:

```{r missing 2}

diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))

```

Like R, ggplot2 subscribes to the philosophy that missing values should never silently go missing. It’s not obvious where you should plot missing values, so ggplot2 doesn’t include them in the plot, but it does warn that they’ve been removed:

```{r missing 3, warning=FALSE}

ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point()

```

To suppress that warning, set na.rm = TRUE:

```{r missing 4}

ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE)

```

Other times you want to understand what makes observations with missing values different to observations with recorded values. For example, in nycflights13::flights, missing values in the dep_time variable indicate that the flight was cancelled. So you might want to compare the scheduled departure times for cancelled and non-cancelled times. You can do this by making a new variable with is.na().

```{r missing 5}

nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)

```

### 7.5 Covariation

It’s common to want to explore the distribution of a continuous variable broken down by a categorical variable, as in the previous frequency polygon. The default appearance of geom_freqpoly() is not that useful for that sort of comparison because the height is given by the count. That means if one of the groups is much smaller than the others, it’s hard to see the differences in shape. For example, let’s explore how the price of a diamond varies with its quality:

```{r covariation 1}

ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

```

It’s hard to see the difference in distribution because the overall counts differ so much:

```{r covariation 2}

ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut))

```

To make the comparison easier we need to swap what is displayed on the y-axis. Instead of displaying count, we’ll display density, which is the count standardised so that the area under each frequency polygon is one.

```{r covariation 3}

ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)

```

Let’s take a look at the distribution of price by cut using geom_boxplot():

```{r covariation 4}

ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()

```

For example, take the class variable in the mpg dataset. You might be interested to know how highway mileage varies across classes:

```{r covariation 5}

ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()

```

To make the trend easier to see, we can reorder class based on the median value of hwy:

```{r covariation 6}

ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))

```

If you have long variable names, geom_boxplot() will work better if you flip it 90°. You can do that with coord_flip().

```{r covariation 7}

ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip()

```

To visualise the covariation between categorical variables, you’ll need to count the number of observations for each combination. One way to do that is to rely on the built-in geom_count():

```{r covariation 8}

ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = color))

```

Another approach is to compute the count with dplyr:

```{r covariation 9}

diamonds %>% 
  count(color, cut)

```

Then visualise with geom_tile() and the fill aesthetic:

```{r covariation 10}

diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))

```

#### Two continuous variables

You’ve already seen one great way to visualise the covariation between two continuous variables: draw a scatterplot with geom_point(). You can see covariation as a pattern in the points. For example, you can see an exponential relationship between the carat size and price of a diamond.

```{r covariation 11}

ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))

```

Scatterplots become less useful as the size of your dataset grows, because points begin to overplot, and pile up into areas of uniform black (as above). You’ve already seen one way to fix the problem: using the alpha aesthetic to add transparency.

```{r covariation 12}

ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = carat, y = price), alpha = 1 / 100)

```

But using transparency can be challenging for very large datasets. Another solution is to use bin. Previously you used geom_histogram() and geom_freqpoly() to bin in one dimension. Now you’ll learn how to use geom_bin2d() and geom_hex() to bin in two dimensions.

geom_bin2d() and geom_hex() divide the coordinate plane into 2d bins and then use a fill color to display how many points fall into each bin. geom_bin2d() creates rectangular bins. geom_hex() creates hexagonal bins. You will need to install the hexbin package to use geom_hex().

```{r covariation 13}

ggplot(data = smaller) +
  geom_bin2d(mapping = aes(x = carat, y = price))

# install.packages("hexbin")
ggplot(data = smaller) +
  geom_hex(mapping = aes(x = carat, y = price))

```

Another option is to bin one continuous variable so it acts like a categorical variable. Then you can use one of the techniques for visualising the combination of a categorical and a continuous variable that you learned about. For example, you could bin carat and then for each group, display a boxplot:

```{r covariation 14}

ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)))

```

Another approach is to display approximately the same number of points in each bin. That’s the job of cut_number():

```{r covariation 15}

ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))

```

### 7.6 Patterns and models

Patterns in your data provide clues about relationships. If a systematic relationship exists between two variables it will appear as a pattern in the data. If you spot a pattern, ask yourself:

A scatterplot of Old Faithful eruption lengths versus the wait time between eruptions shows a pattern: longer wait times are associated with longer eruptions. The scatterplot also displays the two clusters that we noticed above.

```{r patterns 1}

ggplot(data = faithful) + 
  geom_point(mapping = aes(x = eruptions, y = waiting))

```

Models are a tool for extracting patterns out of data. For example, consider the diamonds data. It’s hard to understand the relationship between cut and price, because cut and carat, and carat and price are tightly related. It’s possible to use a model to remove the very strong relationship between price and carat so we can explore the subtleties that remain. The following code fits a model that predicts price from carat and then computes the residuals (the difference between the predicted value and the actual value). The residuals give us a view of the price of the diamond, once the effect of carat has been removed.

```{r patterns 2}

library(modelr)

mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>% 
  add_residuals(mod) %>% 
  mutate(resid = exp(resid))

ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))

```

Once you’ve removed the strong relationship between carat and price, you can see what you expect in the relationship between cut and price: relative to their size, better quality diamonds are more expensive.

```{r patterns 3}

ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))

```

# End
