---
title: "Investigating NASA Voyager Merged Data"
author: "Geirmundur Orri Sigurdsson"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

```{r dependencies, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(data.table)
```

# Functions

## Function for formatting linear model

```{r function format model}
formatmodel <- function(m, eq_digits = 2, r_digits = 3) {
  eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, list(
    a = format(unname(coef(m)[1]), digits = eq_digits),
    b = format(unname(coef(m)[2]), digits = eq_digits),
    r2 = format(summary(m)$r.squared, digits = r_digits)))
  as.character(as.expression(eq));
}
```

# Download Data

## Voyager 1 and 2 daily merged data

Format information see:
* [VOYAGER-1 INTERPLANETARY DATA](https://spdf.gsfc.nasa.gov/pub/data/voyager/voyager1/merged/vy1mgd.txt)
* [VOYAGER-2 INTERPLANETARY DATA](https://spdf.gsfc.nasa.gov/pub/data/voyager/voyager2/merged/vy2mgd.txt)

```{r download trajectory data, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
rdmv1 <- fread("https://spdf.gsfc.nasa.gov/pub/data/voyager/voyager1/merged/voyager1_daily.asc")
rdmv2 <- fread("https://spdf.gsfc.nasa.gov/pub/data/voyager/voyager2/merged/voyager2_daily.asc")
```

# Convert date time

```{r convert time}
dmv1 <- rdmv1 %>% mutate(ts = date_decimal(V1) + days(V2) + hours(V3))
dmv2 <- rdmv2 %>% mutate(ts = date_decimal(V1) + days(V2) + hours(V3))
```

# Select and rename

```{r select and rename}
dmv1 <- dmv1 %>% select(
  ts,             # Date time
  V1HRANG  = V4,  # Spacecraft Heliographic distance (AU)
  V1HELLAT = V5,  # Heliographic Inertial latitude of the Spacecraft (Degrees, +/-90.)
  V1HELLON = V6,  # Heliographic Inertial longitude of the Spacecraft (Degrees, 0-360)
  V1FMAB   = V7,  # Field Magnitude Average |B| (1/N SUM |B|, nT)
  V1MAF    = V8,  # Magnitude of Average Field (sqrt(Bx^2+By^2+Bz^2), nT)
  V1RTNBR  = V9,  # BR RTN-Coordinate System (nanoteslas)
  V1RTNBT  = V10, # BT RTN-Coordinate System (nanoteslas)
  V1RTNBN  = V11, # BN RTN-Coordinate System (nanoteslas)
  V1RTNPFS = V12, # Proton flow speed, RTN (km/s)
  V1RTNTHE = V13, # THETA-elevation angle of flow velocity vector (degrees in RTN-cordinate system)
  V1RTNPHI = V14, # PHI- azimuth angle of flow velocity vector (degrees in RTN-cordinate system)
  V1PD     = V15, # Proton density (n/cc)
  V1PT     = V16, # Proton Temperature (K) calculated from thermal speed width T=60.5*Vth*Vth
  # LECP H in 1/(cm^2 sec ster MeV)
  V1LECP1  = V17, # 0.57-1.78 H MeV
  V1LECP2  = V18, # 3.40-17.6 H Mev
  V1LECP3  = V19, # 22.0-31.0 H MeV
  # CRS H in 1/(cm^2 sec ster MeV)
  V1CRS01  = V20, # 3.000-4.600 MeV
  V1CRS02  = V21, # 4.600-6.200 MeV
  V1CRS03  = V22, # 6.200-7.700 MeV
  V1CRS04  = V23, # 7.700-12.800 MeV
  V1CRS05  = V24, # 12.800-17.900 MeV
  V1CRS06  = V25, # 17.900-30.000 MeV
  V1CRS07  = V26, # 30.000-48.000 MeV
  V1CRS08  = V27, # 48.000-56.000 MeV
  V1CRS09  = V28, # 74.471-83.661 MeV
  V1CRS10  = V29, # 132.834-154.911 MeV
  V1CRS11  = V30, # 154.911-174.866 MeV
  V1CRS12  = V31, # 174.866-187.713 MeV
  V1CRS13  = V32, # 187.713-220.475 MeV
  V1CRS14  = V33, # 220.475-270.050 MeV
  V1CRS16  = V34) # 270.050- 346.034 MeV
dmv2 <- dmv2 %>% select(
  ts,             # Date time
  V2HRANG  = V4,  # Spacecraft Heliographic distance (AU)
  V2HELLAT = V5,  # Heliographic Inertial latitude of the Spacecraft (Degrees, +/-90.)
  V2HELLON = V6,  # Heliographic Inertial longitude of the Spacecraft (Degrees, 0-360)
  V2FMAB   = V7,  # Field Magnitude Average |B| (1/N SUM |B|, nT)
  V2MAF    = V8,  # Magnitude of Average Field (sqrt(Bx^2+By^2+Bz^2), nT)
  V2RTNBR  = V9,  # BR RTN-Coordinate System (nanoteslas)
  V2RTNBT  = V10, # BT RTN-Coordinate System (nanoteslas)
  V2RTNBN  = V11, # BN RTN-Coordinate System (nanoteslas)
  V2RTNBFS = V12, # Proton flow speed, RTN (km/s)
  V2RTNTHE = V13, # THETA-elevation angle of flow velocity vector (degrees in RTN-cordinate system)
  V2RTNPHI = V14, # PHI- azimuth angle of flow velocity vector (degrees in RTN-cordinate system)
  V2PPD    = V15, # Plasma Proton density (n/cc)
  V2PPT    = V16, # Plasma Proton Temperature (K) calculated from thermal speed width T=60.5*Vth*Vth
  # LECP H in 1/(cm^2 sec ster MeV)
  V2LECP1  = V17, # 0.52-1.45 MeV
  V2LECP2  = V18, # 3.04-17.3 Mev
  V2LECP3  = V19, # 22.0-30.0 MeV
  # CRS H in 1/(cm^2 sec ster MeV)
  V2CRS01  = V20, # 3.000-4.600 MeV
  V2CRS02  = V21, # 4.600-6.200 MeV
  V2CRS03  = V22, # 6.200-7.700 MeV
  V2CRS04  = V23, # 7.700-12.800 MeV
  V2CRS05  = V24, # 12.800-17.900 MeV
  V2CRS06  = V25, # 17.900-30.000 MeV
  V2CRS07  = V26, # 30.000-48.000 MeV
  V2CRS08  = V27, # 48.000-56.000 MeV
  V2CRS09  = V28, # 75.861-82.562 MeV
  V2CRS10  = V29, # 130.339 - 154.217 MeV
  V2CRS11  = V30, # 154.217 - 171.338 MeV
  V2CRS12  = V31, # 171.338 - 193.643 MeV
  V2CRS13  = V32, # 193.643 - 208.152 MeV
  V2CRS14  = V33, # 208.152 - 245.690 MeV
  V2CRS15  = V34, # 245.690 - 272.300 MeV
  V2CRS16  = V35, # 272.300 - 344.010 MeV
  V2CRS17  = V36, # 344.010 - 478.623 MeV
  V2CRS18  = V37) # 478.623 - 598.667 MeV
```

# Replace fill in value with NA

```{r replace fill with na}
dmv1 <- na_if(dmv1, 999.99)
dmv1 <- na_if(dmv1, 9999.9)
dmv1 <- na_if(dmv1, 999.999)
dmv1 <- na_if(dmv1, 99.99999)
dmv1 <- na_if(dmv1, 9999999)
dmv1 <- na_if(dmv1, 9.999E5)
dmv2 <- na_if(dmv2, 999.99)
dmv2 <- na_if(dmv2, 9999.9)
dmv2 <- na_if(dmv2, 999.999)
dmv2 <- na_if(dmv2, 99.99999)
dmv2 <- na_if(dmv2, 9999999)
dmv2 <- na_if(dmv2, 9.999E5)
```

# Convert to data tables

```{r convert to data tables}
setDT(dmv1)
setDT(dmv2)
```

# Combine data

## Combine Voyager 1 and 2 merged data into one set by time

```{r combine v1 and v2 by time}
setkey(dmv1, ts)
setkey(dmv2, ts)
dmv <- dmv2[dmv1, roll = "nearest"]
```

## Combine Voyager 1 and 2 merged data into one set by distance from sun

```{r combine v1 and v2 by time}
dmv1d <- dmv1[,-1] %>% filter(!is.na(V1HRANG))
dmv2d <- dmv2[,-1] %>% filter(!is.na(V2HRANG))
setkey(dmv1d, V1HRANG)
setkey(dmv2d, V2HRANG)
dmvd <- dmv1d[dmv2d, roll = "nearest"]
```

# Estimate the speed

From Trajectory data the speed can be estimated 

```{r speed}
modelv1 <- lm(formula = V1HRANG ~ ts, data = dmv1)
modelv2 <- lm(formula = V2HRANG ~ ts, data = dmv2)
coefficientsv1 <- coef(modelv1)
coefficientsv2 <- coef(modelv2)
speedv1 <- coefficientsv1[2] # AU/μs
speedv2 <- coefficientsv2[2] # AU/μs
speedv1 <- speedv1 * 149597870.7 # Convert to km/s
speedv2 <- speedv2 * 149597870.7 # Convert to km/s
speedv1
speedv2
```

# Plots

* Jupiter                         5.2 AU: dashed brown line
* Saturn                          9.6 AU: dashed red line
* Uranus                         19   AU: dashed green line
* Neptune and Kuiper belt starts 30   AU: dashed blue line
* Kuiper belt ends               50   AU: dashed blue line
* Heliosphere upwind             90   AU: dashed golden line

## Distance from sun

```{r plots distance, message=FALSE, warning=FALSE}
pd <- melt(dmv %>% select(date = ts, "voyager 1" = V1HRANG, "voyager 2" = V2HRANG), id = "date")
ggplot(data = pd, mapping = aes(x = date, y = value, color = variable)) +
  geom_point() + geom_smooth(method = 'lm') +
  labs(title = "Voyager 1 and 2 Distance from Sun", x = "Date", y = "Heliographic distance (AU)") +
  geom_hline(yintercept = 5.2, linetype = 'dashed', color = 'brown') +
  geom_hline(yintercept = 9.6, linetype = 'dashed', color = 'red') +
  geom_hline(yintercept = 19, linetype = 'dashed', color = 'green') +
  geom_hline(yintercept = 30, linetype = 'dashed', color = 'blue') +
  geom_hline(yintercept = 50, linetype = 'dashed', color = 'blue') +
  geom_hline(yintercept = 90, linetype = 'dashed', color = 'darkgoldenrod') +
  theme_light()
```

## Plot Voyager 1 and 2 Proton density

```{r plots proton density}
pd <- melt(dmv %>% select(date = ts, "voyager 1" = V1PD, "voyager 2" = V2PPD), id = "date") %>% filter(!is.na(value))
ggplot(data = pd, mapping = aes(x = date, y = value, color = variable)) +
  geom_point() +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Voyager 1 and 2 Proton Density", x = "Date", y = "Density (n/cc)") +
  theme_light()
pd <- melt(dmvd %>% select(distance = V1HRANG, "voyager 1" = V1PD, "voyager 2" = V2PPD), id = "distance") %>% filter(!is.na(value))
ggplot(data = pd, mapping = aes(x = distance, y = value, color = variable)) +
  geom_line() +
  scale_y_continuous(trans = 'log10') +
  labs(title = "Voyager 1 and 2 Proton Density", x = "Distance (AU)", y = "Density (n/cc)") +
  theme_light()
```

