---
title: "Investigating NASA New Horizons SWAP valid sum 2008"
author: "Geirmundur Orri Sigurdsson"
date: "4/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[The R Markdown code is available on GitHub](https://github.com/orri93/analysis/blob/master/etc/astronomy/nasa/inspect-new-horizons-swap-validsum-2008.Rmd)

# Dependencies

```{r dependencies, message=FALSE, warning=FALSE}
library(lubridate)
library(tidyverse)
library(hrbrthemes)
library(viridis)
```

# Functions

## Function for aggregation and formatting linear model

```{r function for aggregation}
aggregatedf <- function(df, n = 5, FUN = mean) {
  aggregate(df,
            by = list(gl(ceiling(nrow(df)/n), n)[1:nrow(df)]),
            FUN = FUN)[-1]
}
```

## Function for formatting linear model

```{r function format model}
formatmodel <- function(m, eq_digits = 2, r_digits = 3) {
  eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, list(
    a = format(unname(coef(m)[1]), digits = eq_digits),
    b = format(unname(coef(m)[2]), digits = eq_digits),
    r2 = format(summary(m)$r.squared, digits = r_digits)))
  as.character(as.expression(eq));
}
```

# Read Data

```{r read data}
rnhvs <- read.csv('../../../tmp/nasa/spdf/new-horizons/swap/validsum/2008/new_horizons_swap_validsum_20081010210700_v1.0.5.csv')
```

# Convert date time

```{r convert time}
nhvs <- rnhvs %>% mutate(ts = as_datetime(time / 1000))
```

# Range

# Time range

```{r time range}
summary(nhvs$ts)
```

# Distance range

```{r distance range}
summary(nhvs$nhhgidr)
```

# Aggregate by Mean

```{r aggregate by mean}
nhvs <- aggregatedf(nhvs, 50)
```

# Estimate the speed

Compare with 12.26 km/s according to Wikipedia

```{r speed}
model <- lm(formula = nhhgidr ~ time, data = nhvs)
coefficients <- coef(model)
speed <- coefficients[2] # AU/μs
speed <- speed * 149597870.7 * 1000 # Convert to km/s
speed
```

# Plots

## Distance from sun

```{r plots distance sun, message=FALSE, warning=FALSE}
speedtext <- sprintf("Estimated speed %f km/s", speed)
modeleq <- formatmodel(model, eq_digits = 3, r_digits = 4)
ggplot(data = nhvs, mapping = aes(x = ts, y = nhhgidr)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  labs(title = "New Horizons SWAP data", subtitle = speedtext, x = "Time", y = "Distance from Sun (AU)") +
  geom_text(data = nhvs[25,], label = modeleq, vjust = -5, parse = TRUE) +
  theme_light()
```

## Density with Temperature and Speed

```{r plots density, warning=FALSE}
nhvs %>% select(ts, n, speed = v) %>% 
  ggplot(mapping = aes(x = ts, y = n, color = speed)) + geom_point() +
  labs(title = "New Horizons SWAP proton density", x = "Time", y = "cm**-3") +
  theme_light()
nhvs %>% select(nhhgidr, n, speed = v) %>%
  ggplot(mapping = aes(x = nhhgidr, y = n, color = speed)) + geom_point() +
  labs(title = "New Horizons SWAP proton density", x = "Distance from Sun (AU)", y = "cm**-3") +
  theme_light()
nhvs %>% select(ts, n, temperature = t) %>% 
  ggplot(mapping = aes(x = ts, y = n, color = temperature)) + geom_point() +
  labs(title = "New Horizons SWAP proton density", x = "Time", y = "cm**-3") +
  theme_light()
nhvs %>% select(nhhgidr, n, temperature = t) %>%
  ggplot(mapping = aes(x = nhhgidr, y = n, color = temperature)) + geom_point() +
  labs(title = "New Horizons SWAP proton density", x = "Distance from Sun (AU)", y = "cm**-3") +
  theme_light()
```

## Speed with Density and Speed

```{r plots speed, warning=FALSE}
nhvs %>% select(ts, v, density = n) %>% 
  ggplot(mapping = aes(x = ts, y = v, color = density)) + geom_point() +
  labs(title = "New Horizons SWAP proton Speed", x = "Time", y = "km/s") +
  theme_light()
nhvs %>% select(nhhgidr, v, density = n) %>%
  ggplot(mapping = aes(x = nhhgidr, y = v, color = density)) + geom_point() +
  labs(title = "New Horizons SWAP proton Speed", x = "Distance from Sun (AU)", y = "km/s") +
  theme_light()
nhvs %>% select(ts, v, temperature = t) %>% 
  ggplot(mapping = aes(x = ts, y = v, color = temperature)) + geom_point() +
  labs(title = "New Horizons SWAP proton Speed", x = "Time", y = "km/s") +
  theme_light()
nhvs %>% select(nhhgidr, v, temperature = t) %>%
  ggplot(mapping = aes(x = nhhgidr, y = v, color = temperature)) + geom_point() +
  labs(title = "New Horizons SWAP proton Speed", x = "Distance from Sun (AU)", y = "km/s") +
  theme_light()
```


## Temperature with Density and Speed

```{r plots temperature, warning=FALSE}
nhvs %>% select(ts, t, density = n) %>% 
  ggplot(mapping = aes(x = ts, y = t, color = density)) + geom_point() +
  labs(title = "New Horizons SWAP proton Temperature", x = "Time", y = "K") +
  theme_light()
nhvs %>% select(nhhgidr, t, density = n) %>%
  ggplot(mapping = aes(x = nhhgidr, y = t, color = density)) + geom_point() +
  labs(title = "New Horizons SWAP proton Temperature", x = "Distance from Sun (AU)", y = "K") +
  theme_light()
nhvs %>% select(ts, t, speed = v) %>% 
  ggplot(mapping = aes(x = ts, y = t, color = speed)) + geom_point() +
  labs(title = "New Horizons SWAP proton Temperature", x = "Time", y = "K") +
  theme_light()
nhvs %>% select(nhhgidr, t, speed = v) %>%
  ggplot(mapping = aes(x = nhhgidr, y = t, color = speed)) + geom_point() +
  labs(title = "New Horizons SWAP proton Temperature", x = "Distance from Sun (AU)", y = "K") +
  theme_light()
```

## Density, Temperature and Speed

```{r plots density pressure, warning=FALSE, message=FALSE}
ggplot(data = nhvs, mapping = aes(x = n, y = t, color = v)) +
  geom_point() +
  labs(title = "New Horizons SWAP protons", x = "Density (cm**-3)", y = "Temperature (K)") +
  theme_light()
ggplot(data = nhvs, mapping = aes(x = v, y = t, color = n)) +
  geom_point() +
  labs(title = "New Horizons SWAP protons", x = "Speed (km/s)", y = "Temperature (K)") +
  theme_light()
ggplot(data = nhvs, mapping = aes(x = n, y = v, color = t)) +
  geom_point() +
  labs(title = "New Horizons SWAP protons", x = "Density (cm**-3)", y = "Speed (km/s)") +
  theme_light()
```
