---
title: "PID tuning comparison"
author: "Geirmundur Orri Sigurdsson"
date: "16 4 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PID Tuning Report

```{r echo = FALSE, warning = FALSE, message = FALSE, results="hide", fig.show="hide", comment=NA}
library(knitr)
library(kableExtra)

processpidformd <- function(datafilepath, iscale = 0.1, yl = -1, xl = -1, ymax = -1, xmax = -1, w = -1) {
  
  require(R.utils)
  require(zoo)
  
  dcsv <- read.csv(datafilepath)
  
  if(w < 0) {
    w = 25
  }
  
  time <- dcsv[[1]]
  status <- dcsv[[2]]
  kp <- max(dcsv[[3]])
  ki <- max(dcsv[[4]])
  kd <- max(dcsv[[5]])
  setpoint <- dcsv[[6]]
  output <- dcsv[[7]]
  temperature <- dcsv[[8]]
  error <- dcsv[[9]]
  integral <- dcsv[[10]]
  derivative <- dcsv[[11]]
  
  scaled_integral <- integral * iscale
  
  data_list <- list(
    "time" = time,
    "kp" = kp,
    "ki" = ki,
    "kd" = kd,
    "setpoint" = setpoint,
    "output" = output,
    "temperature" = temperature,
    "error" = error,
    "integral" = integral,
    "derivative" = derivative,
    "scaled_integral" = scaled_integral)
  
  count <- length(temperature)
  fixed_setpoint <- max(setpoint)
  maximum_temperature <- max(temperature)
  
  mtime = rollmean(time, w)
  mtemperature <- rollmean(temperature, w)
  mcount = length(mtemperature)
  over <- FALSE
  under <- FALSE
  for(i in 1:mcount) {
    if(over && under) {
      break
    } else {
      if(over) {
        if(mtemperature[i] < fixed_setpoint) {
          under <- TRUE
          undertime <- mtime[i]
        }
      } else {
        if(mtemperature[i] > fixed_setpoint) {
          over <- TRUE
          overtime <- mtime[i]
        }
      }
    }
  }
  
  startindex <- -1
  diffcount <- 0
  diffsqsum <- 0
  for(i in 1:count) {
    if(time[i] > undertime) {
      diffcount <- diffcount + 1
      diffsqsum <- diffsqsum + error[i] * error[i]
      if(startindex < 0) {
        startindex = i
      }
    }
  }
  diffsqavr <- diffsqsum / diffcount
  diffsqrt <- sqrt(diffsqavr)
  
  timescale_list <- list("minimum" = min(time), "maximum" = max(time))
  
  calculations_list <- list(
    "timescale" = timescale_list,
    "count" = count,
    "fixed_setpoint" = fixed_setpoint,
    "maximum_temperature" = maximum_temperature,
    "overtime" = overtime,
    "undertime" = undertime,
    "startindex" = startindex,
    "diffcount" = diffcount,
    "diffsqsum" = diffsqsum,
    "diffsqavr" = diffsqavr,
    "diffsqrt" = diffsqrt,
    "peak_error" = maximum_temperature - fixed_setpoint)
  
  if(ymax < 0) {
    ymax = max(
      fixed_setpoint,
      max(output),
      maximum_temperature,
      max(error),
      max(scaled_integral),
      max(derivative))
  }
  if(xmax < 0) {
    xmax <- max(mtime)
  }
  if(yl < 0) {
    yl = ymax - 10
  }
  if(xl < 0) {
    xl = 1000
  }
  limx <- c(0, xmax)
  limy <- c(0, ymax)
    
  
  leg <- c('temperature','output', 'setpoint', 'error', 'integral', 'derivative')
  cols <- c('chocolate', 'blue', 'green', 'brown', 'darkviolet', 'cyan')
  pchs <- c(1,2,3,4,5,6)
  
  labx <- "time"
  laby <- "temperature"
  
  plot_list <- list(
    "ymax" = ymax,
    "xmax" = xmax,
    "yl" = yl,
    "xl" = xl,
    "limx" = limx,
    "limy" = limy,
    "leg" = leg,
    "cols" = cols,
    "pchs" = pchs,
    "labx" = labx,
    "laby" = laby,
    "title"= "PID result",
    "sub" = datafilepath)
  
  result_list <- list(
    "data" = data_list,
    "calculations" = calculations_list,
    "plot" = plot_list)
  
  return(result_list)
}

gentable <- function(pd) {
  return(data.frame(
    Parameter = c("Count", "Setpoint"),
    Value = c(pd$calculations$count, pd$calculations$fixed_setpoint)))
}

tuningtable <- function(pd) {
  return(data.frame(
    Parameter = c("Kp", "Ki", "Kd"),
    Value = c(pd$data$kp, pd$data$ki, pd$data$kd)))
}

calctable <- function(pd) {
  return(data.frame(
    Parameter = c("Maximum Temperature", "Peak Error", "General Error"),
    Value = c(
      pd$calculations$maximum_temperature,
      pd$calculations$peak_error,
      pd$calculations$diffsqrt)))
}

detailtable <- function(pd) {
  return(data.frame(
    Parameter = c(
      "Over time",
      "Under time",
      "Start Index",
      "End Index",
      "General Error Count",
      "Error Squared Sum",
      "Error Squared Avrage"),
    Value = c(
      pd$calculations$overtime,
      pd$calculations$undertime,
      pd$calculations$startindex,
      pd$calculations$count,
      pd$calculations$diffcount,
      pd$calculations$diffsqsum,
      pd$calculations$diffsqavr)))
}

generateplot <- function(pd) {
  plot(
    pd$data$time,
    pd$data$temperature,
    xlab=pd$plot$labx,
    ylab=pd$plot$laby,
    ylim=pd$plot$limy,
    col=pd$plot$cols[0],
    pch=pd$plot$pchs[0])
  par(new=TRUE)
  plot(
    pd$data$time,
    pd$data$output,
    xlab=pd$plot$labx,
    ylab=pd$plot$laby,
    ylim=pd$plot$limy,
    col=pd$plot$cols[1],
    pch=pd$plot$pchs[1])
  par(new=TRUE)
  plot(
    pd$data$time,
    pd$data$setpoint,
    xlab=pd$plot$labx,
    ylab=pd$plot$laby,
    ylim=pd$plot$limy,
    col=pd$plot$cols[2],
    pch=pd$plot$pchs[2])
  par(new=TRUE)
  plot(
    pd$data$time,
    pd$data$error,
    xlab=pd$plot$labx,
    ylab=pd$plot$laby,
    ylim=pd$plot$limy,
    col=pd$plot$cols[3],
    pch=pd$plot$pchs[3])
  par(new=TRUE)
  plot(
    pd$data$time,
    pd$data$scaled_integral,
    xlab=pd$plot$labx,
    ylab=pd$plot$laby,
    ylim=pd$plot$limy,
    col=pd$plot$cols[4],
    pch=pd$plot$pchs[4])
  par(new=TRUE)
  plot(
    pd$data$time,
    pd$data$derivative,
    xlab=pd$plot$labx,
    ylab=pd$plot$laby,
    ylim=pd$plot$limy,
    col=pd$plot$cols[5],
    pch=pd$plot$pchs[5])
  legend(
    x=600,
    y=410,
    legend=pd$plot$leg,
    col=pd$plot$cols,
    pch=pd$plot$pchs)
  title(main = pd$plot$title, sub = pd$plot$suv)
}

creategenerictable <- function(pd) {
  kable(gentable(pd)) %>%
    kable_styling(full_width = F) %>%
    column_spec(1, bold = T, border_right = T) %>%
    column_spec(2, border_right = T)
}

createtuningtable <- function(pd) {
  kable(tuningtable(pd)) %>%
    kable_styling(full_width = F) %>%
    column_spec(1, bold = T, border_right = T) %>%
    column_spec(2, border_right = T)
}

createcalculationstable <- function(pd) {
  kable(calctable(pd)) %>%
    kable_styling(full_width = F) %>%
    column_spec(1, bold = T, border_right = T) %>%
    column_spec(2, border_right = T)}

createdetailtable <- function(pd) {
    kable(detailtable(pd)) %>%
    kable_styling(full_width = F) %>%
    column_spec(1, bold = T, border_right = T) %>%
    column_spec(2, border_right = T)
}

#getwd()

p200413c <- processpidformd("../pid/csv/200413c.csv")
p200414a <- processpidformd("../pid/csv/200414a.csv")
p200414b <- processpidformd("../pid/csv/200414b.csv")
p200414c <- processpidformd("../pid/csv/200414c.csv")
p200415a <- processpidformd("../pid/csv/200415a.csv")
p200415b <- processpidformd("../pid/csv/200415b.csv")

```

## Summary

```{r echo = FALSE, result = 'hide'}
summarytable <- data.frame(
    Run = c(
      "2020-04-13 C",
      "2020-04-14 A",
      "2020-04-14 B",
      "2020-04-14 C",
      "2020-04-15 A",
      "2020-04-15 B"),
    Kp = c(
      p200413c$data$kp,
      p200414a$data$kp,
      p200414b$data$kp,
      p200414c$data$kp,
      p200415a$data$kp,
      p200415b$data$kp),
    Ki = c(
      p200413c$data$ki,
      p200414a$data$ki,
      p200414b$data$ki,
      p200414c$data$ki,
      p200415a$data$ki,
      p200415b$data$ki),
    Kd = c(
      p200413c$data$kd,
      p200414a$data$kd,
      p200414b$data$kd,
      p200414c$data$kd,
      p200415a$data$kd,
      p200415b$data$kd),
    Setpoint = c(
      p200413c$calculations$fixed_setpoint,
      p200414a$calculations$fixed_setpoint,
      p200414b$calculations$fixed_setpoint,
      p200414c$calculations$fixed_setpoint,
      p200415a$calculations$fixed_setpoint,
      p200415b$calculations$fixed_setpoint),
    Peak = c(
      p200413c$calculations$maximum_temperature,
      p200414a$calculations$maximum_temperature,
      p200414b$calculations$maximum_temperature,
      p200414c$calculations$maximum_temperature,
      p200415a$calculations$maximum_temperature,
      p200415b$calculations$maximum_temperature),
     "Peak Error" = c(
      p200413c$calculations$peak_error,
      p200414a$calculations$peak_error,
      p200414b$calculations$peak_error,
      p200414c$calculations$peak_error,
      p200415a$calculations$peak_error,
      p200415b$calculations$peak_error),
    "General Error" = c(
      p200413c$calculations$diffsqrt,
      p200414a$calculations$diffsqrt,
      p200414b$calculations$diffsqrt,
      p200414c$calculations$diffsqrt,
      p200415a$calculations$diffsqrt,
      p200415b$calculations$diffsqrt))
kable(summarytable) %>%
  kable_styling(full_width = F) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, border_right = T) %>%
  column_spec(3, border_right = T) %>%
  column_spec(4, border_right = T) %>%
  column_spec(5, border_right = T) %>%
  column_spec(6, border_right = T) %>%
  column_spec(7, border_right = T) %>%
  column_spec(8, border_right = T)
```

## PID tuning 2020-04-13 C

```{r echo = FALSE, result = 'hide'}
data <- p200413c
```

Time scale is from `r data$calculations$timescale$minimum` to `r data$calculations$timescale$maximum`

The count is `r data$calculations$count`

### General Parameters

```{r echo = FALSE, result = 'asis'}
creategenerictable(data)
```

### Tuning Parameters

```{r echo = FALSE, result = 'asis'}
createtuningtable(data)
```

### Calculated Values

```{r echo = FALSE, result = 'asis'}
createcalculationstable(data)
```


### Calculation Details

```{r echo = FALSE, result = 'asis'}
createdetailtable(data)
```

### Chart

```{r echo=FALSE}
generateplot(data)
```

## PID tuning 2020-04-14 A

```{r echo = FALSE, result = 'hide'}
data <- p200414a
```

Time scale is from `r data$calculations$timescale$minimum` to `r data$calculations$timescale$maximum`

The count is `r data$calculations$count`

### General Parameters

```{r echo = FALSE, result = 'asis'}
creategenerictable(data)
```

### Tuning Parameters

```{r echo = FALSE, result = 'asis'}
createtuningtable(data)
```

### Calculated Values

```{r echo = FALSE, result = 'asis'}
createcalculationstable(data)
```


### Calculation Details

```{r echo = FALSE, result = 'asis'}
createdetailtable(data)
```

### Chart

```{r echo=FALSE}
generateplot(data)
```


## PID tuning 2020-04-14 B

```{r echo = FALSE, result = 'hide'}
data <- p200414b
```

Time scale is from `r data$calculations$timescale$minimum` to `r data$calculations$timescale$maximum`

The count is `r data$calculations$count`

### General Parameters

```{r echo = FALSE, result = 'asis'}
creategenerictable(data)
```

### Tuning Parameters

```{r echo = FALSE, result = 'asis'}
createtuningtable(data)
```

### Calculated Values

```{r echo = FALSE, result = 'asis'}
createcalculationstable(data)
```


### Calculation Details

```{r echo = FALSE, result = 'asis'}
createdetailtable(data)
```

### Chart

```{r echo=FALSE}
generateplot(data)
```


## PID tuning 2020-04-14 C

```{r echo = FALSE, result = 'hide'}
data <- p200414c
```

Time scale is from `r data$calculations$timescale$minimum` to `r data$calculations$timescale$maximum`

The count is `r data$calculations$count`

### General Parameters

```{r echo = FALSE, result = 'asis'}
creategenerictable(data)
```

### Tuning Parameters

```{r echo = FALSE, result = 'asis'}
createtuningtable(data)
```

### Calculated Values

```{r echo = FALSE, result = 'asis'}
createcalculationstable(data)
```


### Calculation Details

```{r echo = FALSE, result = 'asis'}
createdetailtable(data)
```

### Chart

```{r echo=FALSE}
generateplot(data)
```




## PID tuning 2020-04-15 A

```{r echo = FALSE, result = 'hide'}
data <- p200415a
```

Time scale is from `r data$calculations$timescale$minimum` to `r data$calculations$timescale$maximum`

The count is `r data$calculations$count`

### General Parameters

```{r echo = FALSE, result = 'asis'}
creategenerictable(data)
```

### Tuning Parameters

```{r echo = FALSE, result = 'asis'}
createtuningtable(data)
```

### Calculated Values

```{r echo = FALSE, result = 'asis'}
createcalculationstable(data)
```


### Calculation Details

```{r echo = FALSE, result = 'asis'}
createdetailtable(data)
```

### Chart

```{r echo=FALSE}
generateplot(data)
```


## PID tuning 2020-04-15 B

```{r echo = FALSE, result = 'hide'}
data <- p200415b
```

Time scale is from `r data$calculations$timescale$minimum` to `r data$calculations$timescale$maximum`

The count is `r data$calculations$count`

### General Parameters

```{r echo = FALSE, result = 'asis'}
creategenerictable(data)
```

### Tuning Parameters

```{r echo = FALSE, result = 'asis'}
createtuningtable(data)
```

### Calculated Values

```{r echo = FALSE, result = 'asis'}
createcalculationstable(data)
```


### Calculation Details

```{r echo = FALSE, result = 'asis'}
createdetailtable(data)
```

### Chart

```{r echo=FALSE}
generateplot(data)
```