---
title: "R 3D Notebook"
output:
  html_document:
    df_print: paged
---

# Dependencies

```{r dependencies}
library(sphereplot)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

# Prepare the data

We’ll use the iris data set in the following examples:

```{r}
data(iris)
head(iris)
```

```{r}
x <- sep.l <- iris$Sepal.Length
y <- pet.l <- iris$Petal.Length
z <- sep.w <- iris$Sepal.Width
```

# Start and close RGL device

The functions below are used to manage the RGL device:

* rgl.open(): Opens a new device
* rgl.close(): Closes the current device
* rgl.clear(): Clears the current device
* rgl.cur(): Returns the active device ID
* rgl.quit(): Shutdowns the RGL device system

# 3D scatter plot

The function rgl.points() is used to draw a 3D scatter plot:

```{r rgl-points, webgl=TRUE}
rgl.open() # Open a new RGL device
rgl.points(x, y, z, color ="lightgray") # Scatter plot
```

# Change the background and point colors

* The function rgl.bg(color) can be used to setup the background environment of the scene
* The argument color is used in the function rgl.points() to change point colors

```{r}
rgl.open()# Open a new RGL device
rgl.bg(color = "white") # Setup the background color
rgl.points(x, y, z, color = "blue", size = 5) # Scatter plot
```

# Change the shape of points

It’s possible to draw spheres using the functions rgl.spheres() or spheres3d():

```
spheres3d(x, y = NULL, z = NULL, radius = 1, ...)
rgl.spheres(x, y = NULL, z = NULL, r, ...)
```

rgl.spheres() draws spheres with center (x, y, z) and radius r.

* x, y, z : Numeric vector specifying the coordinates for the center of each sphere. The arguments y and z are optional when:
* x is a matrix or a data frame containing at least 3 columns which will be used as the x, y and z coordinates. Ex: rgl.spheres(iris, r = 0.2)
* x is a formula of form zvar ~ xvar + yvar (see ?xyz.coords). Ex: rgl.spheres( z ~ x + y, r = 0.2).
* radius: a vector or single value indicating the radius of spheres
* …: Material properties. See ?rgl.material for details.

```{r}
rgl.open()# Open a new RGL device
rgl.bg(color = "white") # Setup the background color
rgl.spheres(x, y, z, r = 0.2, color = "grey")
```

# rgl_init(): A custom function to initialize RGL device

The function rgl_init() will create a new RGL device if requested or if there is no opened device:

```{r}
#' @param new.device a logical value. If TRUE, creates a new device
#' @param bg the background color of the device
#' @param width the width of the device
rgl_init <- function(new.device = FALSE, bg = "white", width = 640) { 
  if( new.device | rgl.cur() == 0 ) {
    rgl.open()
    par3d(windowRect = 50 + c( 0, 0, width, width ) )
    rgl.bg(color = bg )
  }
  rgl.clear(type = c("shapes", "bboxdeco"))
  rgl.viewpoint(theta = 15, phi = 20, zoom = 0.7)
}
```

