---
title: "Mass of Energy in Batteries"
author: "Geirmundur Orri Sigurdsson"
date: '2023-01-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Energy mass equivalence of a 9V battery

## Calculations

```{r b9v}
c <- 299792458
csquared <- c ^ 2
energy_per_charge <- 9
charge <- 1800
energy <- energy_per_charge * charge
masskg <- energy / csquared
massmicrog <- masskg * 1000000000
```

## Report

A 9V battery delivers `r toString(energy_per_charge)` joules of energy for each coulomb of charge it delivers. Each battery can typically deliver about `r toString(charge)` coulomb of charge before it is drained. The total energy each battery provides over its life is `r toString(energy)` joules.

The mass equivalence for the energy is given by Einstein's Mass–energy equivalence equation
$$
m = E / c^2 = \frac{`r toString(energy)` \, J}{(`r toString(c)` \, m/s)^2} = \frac{`r toString(energy)` \, kg \cdot m^2 \cdot s^{-2}}{`r toString(csquared)` \, m^2 \cdot s^{-2}} = `r toString(masskg)` \, kg
$$

The mass equivalence for the energy of a 9V battery is about `r toString(round(massmicrog, digits = 6))` µg